
```{r include = FALSE}
source("../tools/chunk-options.R")
```

# Vectorisation

```{r, include=FALSE}
# Silently load in the data so the rest of the lesson works
healthData <- read.csv("https://goo.gl/oqQGKF", header=TRUE)
library(ggplot2)
```

<!--sec data-title="Learning Objective" data-id="obj" data-show=true data-collapse=false ces-->

* To understand vectorised operations in R.

<!--endsec-->

<br>

---

**Table of Contents**

<!-- toc -->

<br>

---

Most of R's functions are vectorised, meaning that the function will
operate on all elements of a vector without needing to loop through
and act on each element one at a time. This makes writing code more
concise, easy to read, and less error prone.


```{r}
x <- 1:4
x * 2
```

The multiplication happened to each element of the vector.

We can also add two vectors together:

```{r}
y <- 6:9
x + y
```

Each element of `x` was added to its corresponding element of `y`:

```{r, eval=FALSE}
x:  1  2  3  4
    +  +  +  +
y:  6  7  8  9
---------------
    7  9 11 13
```


<!--sec data-title="Challenge 1" data-id="ch1" data-show=true data-collapse=false ces-->

Let's try this on the `health` column of the `healthData` dataset.

Make a new column in the `healthData` data frame that contains health rounded to the nearest integer. Check the head or tail of the data frame to make sure it worked.

**Hint**: R has a round() function

<!--endsec-->

Comparison operators, logical operators, and many functions are also
vectorised:


**Comparison operators**

```{r}
x > 2
```

**Logical operators** 
```{r}
a <- x > 3  # or, for clarity, a <- (x > 3)
a
```

<!--sec data-title="Tip: Some useful functions forlogical vectors" data-id="tip1" data-show=true data-collapse=true ces-->

`any()` will return `TRUE` if *any* element of a vector is `TRUE` `all()` will return `TRUE` if *all* elements of a vector are `TRUE`

<!--endsec-->

Most functions also operate element-wise on vectors:

**Functions**
```{r}
x <- 1:4
log(x)
```

Vectorised operations work element-wise on matrices:

```{r}
m <- matrix(1:12, nrow=3, ncol=4)
m * -1  
```
 
<!--sec data-title="Tip: Element-wise vs. matrix multiplication" data-id="tip2" data-show=true data-collapse=true ces-->

Very important: the operator `*` gives you element-wise multiplication!
To do matrix multiplication, we need to use the `%*%` operator:
 
```{r}
m %*% matrix(1, nrow=4, ncol=1)
matrix(1:4, nrow=1) %*% matrix(1:4, ncol=1)
```

For more on matrix algebra, see the [Quick-R reference guide](http://www.statmethods.net/advstats/matrix.html)

<!--endsec-->

<!--sec data-title="Challenge 2" data-id="ch2" data-show=true data-collapse=false ces-->

Given the following matrix:

```{r}
m <- matrix(1:12, nrow=3, ncol=4)
m
```

Write down what you think will happen when you run:

1. `m ^ -1`
2. `m * c(1, 0, -1)`
3. `m > c(0, 20)`
4. `m * c(1, 0, -1, 2)`

Did you get the output you expected? If not, ask a helper!

<!--endsec-->

<!--sec data-title="Challenge 3" data-id="ch3" data-show=true data-collapse=false ces-->

We're interested in looking at the sum of the following sequence of fractions:

```{r, eval=FALSE}
x = 1/(1^2) + 1/(2^2) + 1/(3^2) + ... + 1/(n^2)
```

This would be tedious to type out, and impossible for high values of
n.  Use vectorisation to compute x when n=100. What is the sum when n=10,000?

<!--endsec-->

<br>

---

## Challenge solutions

<!--sec data-title="Solution to Challenge 1" data-id="ch1sol" data-show=true data-collapse=true ces-->

Let's try this on the `health` column of the `healthData` dataset.

Make a new column in the `healthData` data frame that contains health rounded to the nearest integer. Check the head or tail of the data frame to make sure it worked.

Hint: R has a round() function

```{r}
healthData$healthInteger <- round(healthData$health)
head(healthData)
```

<!--endsec-->

<!--sec data-title="Solution to Challenge 2" data-id="ch2sol" data-show=true data-collapse=true ces-->

Given the following matrix:

```{r}
m <- matrix(1:12, nrow=3, ncol=4)
m
```

Write down what you think will happen when you run:

1. `m ^ -1`
    ```{r, echo=FALSE}
    m ^ -1
    ```

2. `m * c(1, 0, -1)`
    ```{r, echo=FALSE}
    m * c(1, 0, -1)
    ```

3. `m > c(0, 20)`
    ```{r, echo=FALSE}
    m > c(0, 20)
    ```

<!--endsec-->

<!--sec data-title="Solution to Challenge 3" data-id="ch3sol" data-show=true data-collapse=true ces-->

We're interested in looking at the sum of the following sequence of fractions:

```{r, eval=FALSE}
x = 1/(1^2) + 1/(2^2) + 1/(3^2) + ... + 1/(n^2)
```

This would be tedious to type out, and impossible for high values of n. Can you use vectorisation to compute x, when n=100? How about when n=10,000?

```{r}
sum(1/(1:100)^2)
sum(1/(1:1e04)^2)
n <- 10000
sum(1/(1:n)^2)
```

We can also obtain the same results using a function:
```{r}
inverse_sum_of_squares <- function(n) {
   sum(1/(1:n)^2)
}
inverse_sum_of_squares(100)
inverse_sum_of_squares(10000)
n <- 10000
inverse_sum_of_squares(n)
```

<!--endsec-->
